---
title: "Lab Assignment 3 Answer Key"
author: "Peter James"
date: "2025-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(sf)
library(MapGAM)
library(tidyverse)
library(tidycensus)
library(flextable)
library(RColorBrewer)
library(tmap)
```

Q1. Bring in CAdata
```{r}
data(CAdata)
ca_pts <- CAdata
ca_proj <- "+proj=lcc +lat_1=40 +lat_2=41.66666666666666 
             +lat_0=39.33333333333334 +lon_0=-122 +x_0=2000000 
             +y_0=500000.0000000002 +ellps=GRS80 
             +datum=NAD83 +units=m +no_defs"

ca_pts <- st_as_sf(CAdata, coords=c("X","Y"), crs=ca_proj)
```

Q2. Use Tidycensus to get 2019-2023 estimates of Census tract median home income
```{r}
ca.tract.medinc <- get_acs(geography = "tract", 
              year = 2023,
              variables = c(medinc = "B19013_001"), 
              state = "CA",
              output = "wide",
              survey = "acs5",
              geometry = TRUE,
              cb = FALSE)
```

Q3. Check projections
```{r}
st_crs(ca_pts)==st_crs(ca.tract.medinc)
```

Q4. Reproject CAdata to match Census data
```{r}
ca_pts_proj<-st_transform(ca_pts,st_crs(ca.tract.medinc))

st_crs(ca_pts_proj)==st_crs(ca.tract.medinc)
```

Q5. Overlap map of CAdata with Census data

```{r}
tmap_mode("plot")
ca.inc.map <- tm_shape(ca.tract.medinc) +
  tm_polygons(col = "medincE", style = "quantile",palette = "Reds",
              title = "Median Income") +
      tm_shape(ca_pts_proj) +
        tm_dots(size=0.25, alpha=0.8, col="blue")
ca.inc.map


#Run Subset so it renders in view mode
#tmap_mode("view")
#ca.tract.medinc_small <- ca.tract.medinc[1:2500, ]
#ca_pts_proj_small <- ca_pts_proj[1:2500, ]

#tm_shape(ca.tract.medinc_small) +
#  tm_polygons(col = "medincE", style = "quantile",palette = "Reds",
#              title = "Median Income") +
#  tm_shape(ca_pts_proj_small) +
#  tm_dots(size=0.5, alpha=0.8, col="blue")
```

Q6. Spatially join cancer data to census data
```{r, eval=TRUE, warning=FALSE, message=FALSE}
## Spatially join the cancer point data to the walkability polygon data
census_cancer = st_join(ca_pts_proj, ca.tract.medinc) 

## Take a look at a summary of the values
summary(census_cancer$medincE)
```

# Drop missing values
Alternate approach
census_cancer_nomiss <- census_cancer %>%
    subset(!is.na(medincE))
```{r}
census_cancer_nomiss <- census_cancer %>% drop_na(medincE) 
summary(census_cancer_nomiss$medincE)
```

Q8. Histogram
```{r}
census_cancer_nomiss %>%
  ggplot() + 
  geom_histogram(mapping = aes(x=medincE)) 

```


Q9. Quartiles

```{r}
census_cancer_nomiss <- census_cancer_nomiss %>%
  mutate(medinc_quartile = ntile(medincE, 4))

glimpse(census_cancer_nomiss)
```

Q10. Tables by event
```{r}
## Create a contingency table of event by walk_quartile
tab <- table(census_cancer_nomiss$medinc_quartile, census_cancer_nomiss$event)
tab
```

\

Hmmm, that's interesting, but let's look at this by percentages instead.
```{r}
## Convert to percentages by column
tab_col_perc <- prop.table(tab, margin = 2) * 100
round(tab_col_perc, 1)
```

Q11. Run Chi-squared test
```{r}
## Chi-squared test
chisq.test(tab)
```

\

OK, how do we interpret this? Our null hypothesis is that there is no association between mortality at end of follow-up and increasing quartile of median income. Our alternative hypothesis is that there is an association between mortality at end of follow-up and increasing quartile of median income. We use a two-sided chi-squared test with alpha=0.05. Assuming no sources of bias and that the null hypothesis is true, the probability of observing increases in mortality at end of follow-up with increasing quartiles of walkability as or more extreme as those produced in these data is 0.2309. Since p>0.05, we fail to reject the null hypothesis and conclude that median income is not associated with mortality at end of follow-up (under the assumptions stated above). In other words, we don't see a relationship between census tract median income exposure and our outcome (dying over followup).

