---
title: "Lab Assignment 4 Answer Key"
author: "Peter James"
date: "2025-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(sf)
library(MapGAM)
library(tidyverse)
library(flextable)
library(RColorBrewer)
library(tmap)
library(terra)
```

Q1. Bring in CAdata
```{r}
data(CAdata)
ca_pts <- CAdata
ca_proj <- "+proj=lcc +lat_1=40 +lat_2=41.66666666666666 
             +lat_0=39.33333333333334 +lon_0=-122 +x_0=2000000 
             +y_0=500000.0000000002 +ellps=GRS80 
             +datum=NAD83 +units=m +no_defs"

ca_pts <- st_as_sf(CAdata, coords=c("X","Y"), crs=ca_proj)
```

Q2. Bring in raster data
```{r}
url <- "https://github.com/pjames-ucdavis/SPH215/raw/main/LST_med_ca.tif"
download.file(url, destfile = "LST_med_ca.tif", mode = "wb")
lst_rast = rast("LST_med_ca.tif")

lst_rast
```

Q3. Check projections
```{r}
st_crs(ca_pts)==st_crs(lst_rast)
```

Q4. Reproject CAdata to match LST data
```{r}
ca_pts_proj<-st_transform(ca_pts,st_crs(lst_rast))

st_crs(ca_pts_proj)==st_crs(lst_rast)
```

Q5. Overlap map of CAdata with LST data

```{r}
ca.lst.map <- tm_shape(lst_rast) +
  tm_raster(style = "cont") +
      tm_shape(ca_pts_proj) +
        tm_dots(size=0.25, alpha=0.8, col="red")
ca.lst.map
```

Q6. Extract light at night data to cancer data
```{r}

lst_cancer = data.frame(ca_pts_proj,terra::extract(lst_rast, ca_pts_proj))
glimpse(lst_cancer) 
## Take a look at a summary of the values
summary(lst_cancer)
```


Q7. Drop missing values
Alternate approach
lst_cancer_nomiss <- lst_cancer %>%
    subset(!is.na(LST_Day_1km_median))
    
```{r}
lst_cancer_nomiss <- lst_cancer %>% drop_na(LST_Day_1km_median) 

## Take a look at a summary of the values
summary(lst_cancer_nomiss$LST_Day_1km_median)
glimpse(lst_cancer_nomiss)
```

Q8. Histogram
```{r}
lst_cancer_nomiss %>%
  ggplot() + 
  geom_histogram(mapping = aes(x=LST_Day_1km_median)) 

```


Q9. Quartiles

```{r}
lst_cancer_nomiss <- lst_cancer_nomiss %>%
  mutate(lst_quartile = ntile(LST_Day_1km_median, 4))

glimpse(lst_cancer_nomiss)
```

Q10. Tables by event
```{r}
## Create a contingency table of event by walk_quartile
tab <- table(lst_cancer_nomiss$lst_quartile, lst_cancer_nomiss$event)
tab
```

\

Hmmm, that's interesting, but let's look at this by percentages instead.
```{r}
## Convert to percentages by column
tab_col_perc <- prop.table(tab, margin = 2) * 100
round(tab_col_perc, 1)
```

Q11. Run Chi-squared test
```{r}
## Chi-squared test
chisq.test(tab)
```

\

OK, how do we interpret this? Our null hypothesis is that there is no association between increasing quartile of light at night and mortality at end of follow-up. Our alternative hypothesis is that there is an association between increasing quartile of light at night and mortality at end of follow-up. We use a two-sided chi-squared test with alpha=0.05. Assuming no sources of bias and that the null hypothesis is true, the probability of observing increases in mortality at end of follow-up with increasing quartiles of light at night as or more extreme as those produced in these data is 0.004391. Since p<0.05, we  reject the null hypothesis and conclude that light at night exposure is associated with mortality at end of follow-up (under the assumptions stated above). In other words, we DO see a relationship between light at night exposure and our outcome (dying over followup).

