---
title: "Lab 2 Assignment"
author: "Peter James"
date: "2025-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We’ll bring in the file NDVI_rast.tif. The file contains normalized difference vegetation index (NDVI) data for the Bay Area. These data are taken from Landsat satellite data that I downloaded from Google Earth Engine. We use the function `rast()` to bring in data in raster form, then take a look at the dataset.

```{r}
url <- "https://github.com/pjames-ucdavis/SPH215/raw/main/LST_med_ca.tif"
download.file(url, destfile = "LST_med_ca.tif", mode = "wb")
LST_raster = rast("LST_med_ca.tif")

## Get summary of raster data
LST_raster
```

\

Does it have a CRS?
```{r}

## Check CRS
st_crs(LST_raster)

```

\

OK we have what looks like a raster. We see our resolution and our extent, and we have a CRS. Nice! Shall we plot this?

```{r}
## Plot the raster on a map
tmap_mode("plot")
LST_map = tm_shape(LST_raster) +
  tm_raster(style = "cont") +
  tm_legend(outside = TRUE)
LST_map
```
