---
title: "Lab 1 Assignment"
author: "Peter James"
date: "2025-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Packages
```{r}
library(sf)
library(tidyverse)
library(tidygeocoder)
library(MapGAM)
library(tidycensus)
library(flextable)
library(tmap)
```

\

# Geocoding

Question 1. Read in In N Out Burger

glimpse(innout)
```{r}
download.file("https://raw.githubusercontent.com/pjames-ucdavis/SPH215/refs/heads/main/innout_final.csv", "innout_final.csv", mode = "wb")
in_n_out <- read_csv("innout_final.csv")
```

\

Question 2. Glimpse at the data. Looks like *store address* is the variable to geocode!
```{r}
glimpse(in_n_out)
```

\

Question 3. Geocode the data using "arcgis" geocoder.
```{r}
innout_geo      <- geocode(in_n_out, address = "store address",
                          method = "arcgis")
```

\

Question 4. Looks like there's no missingness! No NAs!
```{r}
summary(innout_geo$lat)
summary(innout_geo$long)
```

\

Question 5. We use st_as_sf to make the dataset spatial and we use the crs of 4326
```{r}
innout_pts <- st_as_sf(innout_geo, coords=c("long","lat"), crs=4326)
```

\

Question 6. Map the addreses in red below!
```{r}
tmap_mode("view")
innout_map = tm_shape(innout_pts) + tm_dots(col = "red", size = 0.3, alpha = 0.5)
innout_map
```


Question 7. Map the addreses in blue below!
```{r}
tmap_mode("view")
innout_map_blue = tm_shape(innout_pts) + tm_dots(col = "blue", size = 0.3, alpha = 0.5)
innout_map_blue
```

# Map Point Health Data

Question 8. Import CAdata dataset
```{r}
#Load CAdata dataset from MapGAM package
data(CAdata)
ca_pts <- CAdata
summary(ca_pts)
ca_pts <- st_as_sf(CAdata, coords=c("X","Y"))

```

\

Question 9. Add CRS
```{r}

#Load the projection into an object called ca_proj
ca_proj <- "+proj=lcc +lat_1=40 +lat_2=41.66666666666666 
             +lat_0=39.33333333333334 +lon_0=-122 +x_0=2000000 
             +y_0=500000.0000000002 +ellps=GRS80 
             +datum=NAD83 +units=m +no_defs"

#Set CRS
ca_pts_crs <- st_set_crs(ca_pts, ca_proj)
```

\

Question 10. Create a map of the CAdata dataset with points color coded by quintiles of age.
```{r}
cancer_map_age = tm_shape(ca_pts_crs) + tm_dots(col = "AGE", size = 0.3, style = "quantile")
cancer_map_age
```

\

Question 11. Map by insurance status
```{r}
cancer_map_ins = tm_shape(ca_pts_crs) + tm_dots(col = "INS", size = 0.3, style = "cat")
cancer_map_ins
```

\

# Census Data

Question 12. Use the tidycensus package to download American Community Survey Five year estimates from 2018-2022 in California at the Census tract level for the following variable
```{r}
ca <- get_acs(geography = "tract", 
              year = 2022,
              variables = c(medinc = "B19013_001"), 
              state = "CA",
              survey = "acs5",
              output = "wide")

```
\

Question 13.	Take a glimpse at your data.
```{r}
glimpse(ca)
```

\


Question 14.	Give us the mean median household income across tracts in California.
```{r}
ca %>%
  summarize(Mean = mean(medincE))
```

\

Question 15.	Make a table showing the mean, median, and standard deviation of median household income across tract in California, and make it look fancy!
```{r}
summarytable <- ca %>%
  summarize(Mean = mean(medincE),
            Median = median(medincE),
            SD = sd(medincE)
            )
my_table <- flextable(summarytable)
my_table
```

\

Question 16.	Create a histogram of median household income across tracts in California. Make sure it has a label.
```{r}
ca %>%
  ggplot() + 
  geom_histogram(mapping = aes(x=medincE)) +
  xlab("Median household income")
```

\

Question 17.	OK now give me a boxplot! And I want a nice label for this too!
```{r}
cacounty %>%
  ggplot() +
    geom_boxplot(mapping = aes(y = medincE)) +
    ylab("Median household income")
```

